---
title: "Assignment 3: Data Exploration"
author: "Marisa Fajardo, Section #3"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}

```

```{r datasets}
insects<- read.csv("../Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE) #creating name for insects data set
litter<- read.csv("../Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE) #creating name for litter data set



```
## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicologoy of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We might want to know how effective these insecticides are for certain insect species, and we would want to know how lethal these toxins are for beneficial insect species such as pollinators. We would also want to know if insects are gaining resistants to insecticides, thus making them useless. 

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: Wood litter and debris on the forest floor adds a great amount of nutrients to the soil and aids in keeping in moisture. Woody debris also provides habitat for small organisms. Studies on the amount of biomass on forest floors can be indicative of the nutrients in a forest ecosystem and biodiversity within a forest. 

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Litter is collected in plots of varying sizes based on canopy cover
*Litter trap placement in plots are be randomized in plots with >50% woody vegetation cover
*in sites with <50% woody vegetation cover, litter trap placement will be targeted so that traps are placed underneath vegetation to catch litter



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}

dim(insects) #checking dimensions of insects

```


6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(insects$Effect) #viewing summary of Effect column


```

> Answer:  The most common effects that are studied are population, followed by mortality, then behavior. Population would be of importance since we would want to identify species in an area. Mortality is also important to study, since it will show if insect death occurs in response to insecticides. Behavior is also important since it details responses of insects to the insecticides. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(insects$Species.Common.Name) #viewing summary of species common name column

```

> Answer: The 6 most common insects are: Honey bee(667), parasitic wasp(285), buff tailed honeybee(183),  Carniolan Honey Bee(152), Italian honeybee(113). These species are all pollinators. They may be of interest since they are essential to polinating plants and crops, therefore we would not want them to be affected by insecticides. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}
class(insects$Conc.1..Author.) #viewing class of conc 1 author column
```

> Answer: The class is character. This is not numeric because the data is categorical. 



## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 
```{r}
library(ggplot2)
ggplot(insects) +
  geom_freqpoly(aes(x = Publication.Year), bins = 50) #creating simple line plot of number of studies per publication year


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(insects) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) +

  theme(legend.position = "top") #creating line plot that color codes test locations from the data 
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: 
The most common test locations are in the lab. These test locations frequencies ebb and flow, but overall increase over time, peaking at around 2014, then lowering again towards 2020. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(insects, aes(x = Endpoint)) +
  geom_bar() #creating bar graph to view counts of each endpoint type

which.max(table(insects$Endpoint)) #checking the most common endpoint 
```

> Answer: NOEL is the most common endpoint. It is a no observable effect level that refers to the highest concentration that would produce effects that are not significantly different from the responses of the controls. The second most common endpoint is LOEL(i believe that is what the graph says, but it is difficult to read since the endpoint names are on top of each other). LOEL is a lowest observable effect level that refers to the lowest concentration that that produces effects that are significantly different from the responses of the controls. 
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(litter$collectDate) #viewing class of collection date
View(litter$collectDate)


litter$collectDate <- as.Date(litter$collectDate) #making this data recognizable as a date
class(litter$collectDate) #checking class of data 

unique(litter$collectDate)#checking collection dates
```
collect Date is classified as a character, not a date, so I changed it to a date. 

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(litter$plotID) #Looking at counts under plot id column
summary(litter$plotID) #comparing to summary
```

> Answer: 12 plots were sampled at niwot ridge. The summary function gives an overview of the data frame, while the unique function allows you to identify specifics from the data frame. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(litter, aes(x = functionalGroup)) +
  geom_bar() #creating bar graph to view counts of each functional group type


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) #creating box plot of drymass and functional group

ggplot(litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass), 
              draw_quantiles = c(0.25, 0.5, 0.75)) #creating violin plot of drymass and functional group

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: In the violin plot, the range of dryMass is too big, which causes the visualization of the functional group to become squished, making it difficult to read the data. The box plots are much easier to read and are a better data visualization in this case. 

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles seem to have the highest biomass.